---
title: "VB_R_assignment"
autor: Veeraya Bamrung
date: "2026-03-15"
output:
  pdf_document:
    toc: true
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(readr)
library(tidyverse)
library(ggplot2)
library(tibble)
library(dplyr)
```

# Part I. Data Inspection and Processing
## Input and output paths

```{r}
dir.create("02_output_files", showWarnings = FALSE)

geno_file <- "00_data_files/fang_et_al_genotypes.txt"
pos_file  <- "00_data_files/snp_position.txt"
out_dir   <- "02_output_files"

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(out_dir, "maize"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(out_dir, "teosinte"), showWarnings = FALSE, recursive = TRUE)
```

## Read input files

```{r}
geno <- readr::read_tsv(
  geno_file,
  col_types = cols(.default = col_character()),
  progress = FALSE
)

pos_raw <- readr::read_tsv(
  pos_file,
  col_types = cols(.default = col_character()),
  progress = FALSE
)
```

## Basic inspection
```{r}
cat("Genotype file:\n")
cat("  Rows:", nrow(geno), "\n")
cat("  Columns:", ncol(geno), "\n")
cat("  Size (bytes):", file.info(geno_file)$size, "\n\n")

cat("Position file:\n")
cat("  Rows:", nrow(pos_raw), "\n")
cat("  Columns:", ncol(pos_raw), "\n")
cat("  Size (bytes):", file.info(pos_file)$size, "\n\n")
```

## Preview files geno = genotype file, pos_raw = snp_position file
```{r}
head(geno)
tail(geno)
names(geno)
head(pos_raw)
tail(pos_raw)
names(pos_raw)
```

## Standardize key columns
* The first column of the genotype file is used as the sample identifier, and the third column is used as the group variable. From the SNP position file, I keep only the SNP ID, chromosome, and position columns. Output is a count of intersect of SNP ID.
```{r}
sample_col <- names(geno)[1]
group_col  <- names(geno)[3]

pos <- pos_raw %>%
  transmute(
    SNP_ID     = .data[[names(pos_raw)[1]]],
    Chromosome = .data[[names(pos_raw)[3]]],
    Position   = .data[[names(pos_raw)[4]]]
  )

# Check intersect of two files (SNP_ID)

snp_cols <- intersect(names(geno), pos$SNP_ID)

# Error check

if (length(snp_cols) == 0) {
  stop("No overlapping SNP columns were found between genotype data and position data.")
}

# Recheck new variable
length(snp_cols)
```

## Define maize and teosinte groups
```{r}
maize_groups    <- c("ZMMIL", "ZMMLR", "ZMMMR")
teosinte_groups <- c("ZMPBA", "ZMPIL", "ZMPJA")
```

## Transpose the genotype data, standardize missing values, classify SNP position records, and write the final output files.

```{r}

# Convert genotype rows into SNP rows

make_genotype_by_snp <- function(geno_df, groups, sample_col, group_col, snp_cols) {
  sub_df <- geno_df %>%
    filter(.data[[group_col]] %in% groups)
  
  if (nrow(sub_df) == 0) {
    stop("No samples found for groups: ", paste(groups, collapse = ", "))
  }
  
  sample_ids <- make.unique(sub_df[[sample_col]])
  
  geno_matrix <- sub_df %>%
    select(all_of(snp_cols)) %>%
    as.matrix()
  
  geno_t <- as_tibble(t(geno_matrix), .name_repair = "minimal")
  names(geno_t) <- sample_ids
  
  geno_t %>%
    mutate(SNP_ID = snp_cols, .before = 1)
}

# Replace missing genotype values for output files

replace_missing_symbol <- function(df, symbol) {
  genotype_cols <- setdiff(names(df), c("SNP_ID", "Chromosome", "Position"))
  
  df %>%
    mutate(
      across(
        all_of(genotype_cols),
        ~ case_when(
          is.na(.x) ~ symbol,
          str_trim(.x) == "" ~ symbol,
          .x %in% c("NN", "N", "NA", "na","?/?") ~ symbol,
          TRUE ~ .x
        )
      )
    )
}

# Join positions to genotype data and classify known SNPs

prepare_joined_data <- function(geno_snp_df, pos_df) {
  
  pos_clean <- pos_df %>%
    mutate(
      chr_num = suppressWarnings(as.integer(Chromosome)),
      pos_num = suppressWarnings(as.numeric(Position))
    )
  
  multiple_ids <- pos_clean %>%
    count(SNP_ID, name = "n_rows") %>%
    filter(n_rows > 1) %>%
    pull(SNP_ID)
  
  unknown_ids <- pos_clean %>%
    filter(!(SNP_ID %in% multiple_ids)) %>%
    filter(is.na(chr_num) | is.na(pos_num) | !(chr_num %in% 1:10)) %>%
    pull(SNP_ID) %>%
    unique()
  
  known_ids <- pos_clean %>%
    filter(!(SNP_ID %in% multiple_ids)) %>%
    filter(!(SNP_ID %in% unknown_ids)) %>%
    pull(SNP_ID) %>%
    unique()
  
  joined <- pos_clean %>%
    left_join(geno_snp_df, by = "SNP_ID") %>%
    select(SNP_ID, Chromosome, Position, everything(), -chr_num, -pos_num)
  
  list(
    known    = joined %>% filter(SNP_ID %in% known_ids),
    unknown  = joined %>% filter(SNP_ID %in% unknown_ids),
    multiple = joined %>% filter(SNP_ID %in% multiple_ids)
  )
}

# Write required output files

write_output_files <- function(data_list, out_subdir, prefix) {
  
  known <- data_list$known
  
  for (chr in 1:10) {
    chr_data <- known %>%
      filter(Chromosome == as.character(chr))
    
    asc <- chr_data %>%
      mutate(Position_numeric = suppressWarnings(as.numeric(Position))) %>%
      arrange(Position_numeric) %>%
      select(-Position_numeric) %>%
      replace_missing_symbol("?")
    
    desc <- chr_data %>%
      mutate(Position_numeric = suppressWarnings(as.numeric(Position))) %>%
      arrange(desc(Position_numeric)) %>%
      select(-Position_numeric) %>%
      replace_missing_symbol("-")
    
    asc_file  <- file.path(out_subdir, paste0(prefix, "_chr", chr, "_ascending.txt"))
    desc_file <- file.path(out_subdir, paste0(prefix, "_chr", chr, "_descending.txt"))
    
    readr::write_tsv(asc, asc_file, na = "")
    readr::write_tsv(desc, desc_file, na = "")
  }
}
```

## Build maize dataset
```{r}
maize_snp <- make_genotype_by_snp(
  geno_df    = geno,
  groups     = maize_groups,
  sample_col = sample_col,
  group_col  = group_col,
  snp_cols   = snp_cols
)

maize_data <- prepare_joined_data(maize_snp, pos)

write_output_files(
  data_list  = maize_data,
  out_subdir = file.path(out_dir, "maize"),
  prefix     = "maize"
)
```

## Build teosinte dataset
```{r}
teosinte_snp <- make_genotype_by_snp(
  geno_df    = geno,
  groups     = teosinte_groups,
  sample_col = sample_col,
  group_col  = group_col,
  snp_cols   = snp_cols
)

teosinte_data <- prepare_joined_data(teosinte_snp, pos)

write_output_files(
  data_list  = teosinte_data,
  out_subdir = file.path(out_dir, "teosinte"),
  prefix     = "teosinte"
)
```

## Check output summary
* The assignment requires 40 output files total: 20 for maize and 20 for teosinte.
```{r}
all_output_files <- list.files(out_dir, recursive = TRUE, full.names = TRUE)

cat("=== OUTPUT SUMMARY ===\n")
cat("Maize files:", length(list.files(file.path(out_dir, "maize"))), "\n")
cat("Teosinte files:", length(list.files(file.path(out_dir, "teosinte"))), "\n\n")
cat("Output directory:\n", normalizePath(out_dir), "\n")
```

# Part II. Visualization
## Prepare data for plotting
* For the visualization step, I use only the known SNP positions and convert the wide genotype tables into long format. Each genotype is classified as homozygous, heterozygous, or missing.

```{r}
# Create new directory
dir.create(file.path(out_dir, "figures"), showWarnings = FALSE, recursive = TRUE)

# Convert wide genotype table to long tidy format

maize_joined <- maize_data$known
teosinte_joined <- teosinte_data$known

make_long_genotype <- function(df, group_label) {
  
  genotype_cols <- setdiff(names(df), c("SNP_ID", "Chromosome", "Position"))
  
  df %>%
    pivot_longer(
      cols = all_of(genotype_cols),
      names_to = "SampleID",
      values_to = "Genotype"
    ) %>%
    mutate(
      Group = group_label,
      Genotype = str_squish(as.character(Genotype)),
      State = case_when(
        is.na(Genotype) | Genotype == "" | Genotype %in% c("NN", "N/N", "NA", "N", "?/?") ~ "Missing",
        str_remove_all(Genotype, "/") %in% c("AA", "CC", "GG", "TT") ~ "Homozygous",
        TRUE ~ "Heterozygous"
      )
    )
}

maize_long <- make_long_genotype(maize_joined, "Maize")
teosinte_long <- make_long_genotype(teosinte_joined, "Teosinte")

all_long <- bind_rows(maize_long, teosinte_long) %>%
  mutate(
    Group = factor(Group, levels = c("Maize", "Teosinte")),
    State = factor(State, levels = c("Homozygous", "Heterozygous", "Missing")),
    Chromosome = factor(Chromosome, levels = as.character(1:10))
  )
```

## Plot. I: SNP counts by chromosome

* This plot shows how SNPs are distributed across chromosomes in maize and teosinte. Because the long dataset contains one row per SNP per sample, I keep only distinct SNP IDs before counting.

```{r}
# Make chr into number order
snp_count_plot_data <- all_long %>%
  mutate(Chromosome = factor(Chromosome, levels = 1:10))
# Make a graph
ggplot(snp_count_plot_data, aes(x = Chromosome, fill = Group)) +
  geom_bar(position = "dodge") +
  labs(
    title = "SNP counts by chromosome",
    x = "Chromosome",
    y = "Number of SNPs"
  ) +
  theme_minimal()

# Save Plot
ggsave(
  file.path(out_dir, "figures", "plot_snp_counts_by_chromosome.png"),
  width = 8, height = 5, dpi = 300
)
```

## Plot. II: Proportion of homozygous, heterozygous, and missing sites by group

* This plot summarizes genotype composition at the group level

```{r}
group_plot_data <- all_long %>%
  count(Group, State) %>%
  complete(Group, State, fill = list(n = 0))

ggplot(group_plot_data, aes(x = Group, y = n, fill = State)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of homozygous, heterozygous, and missing sites by group",
    x = "Group",
    y = "Proportion"
  ) +
  theme_minimal()

ggsave(
  file.path(out_dir, "figures", "plot_state_proportion_by_group.png"),
  width = 7, height = 5, dpi = 300
)
```

## Plot. III: My own visualization: proportion of heterozygous sites by chromosome

* As an additional visualization, I plotted the proportion of heterozygous sites by chromosome for maize and teosinte.

```{r}
chrom_het_data <- all_long %>%
  filter(State != "Missing") %>%
  group_by(Group, Chromosome) %>%
  summarise(
    heterozygous_prop = mean(State == "Heterozygous"),
    .groups = "drop"
  )

ggplot(chrom_het_data, aes(x = Chromosome, y = heterozygous_prop, fill = Group)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of heterozygous sites by chromosome",
    x = "Chromosome",
    y = "Heterozygous proportion"
  ) +
  theme_minimal()

ggsave(
  file.path(out_dir, "figures", "plot_heterozygosity_by_chromosome.png"),
  width = 8, height = 5, dpi = 300
)

```



