---
title: "R_Assignment_review_of_Veeraya_for_Valerie"
author: "Veeraya Bamrung"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(readr)
library(tidyverse)
```

# Review overall comments (This review conducted on 10:38 am on March,16th)
* For code-runall, overall, the code run without encountering bug or any issues from R.
* Only the warning receive is _Warning: NAs introduced by coercion_
* All the code covers all assignment part I and part II of the assignment
* Overall is very well done code with very fined details on each sections!

## Review on part I
This include comments from part I assignment

### Data Inspection part
* No issues and looks nice!
* I learn more commands from the part

### Data seperation of mazie and teosinte
* The code is understable and well organized
* The output files looks great and fit the assignment requirement

## Review on part II

### SNPs per chromosome
* It is slightly from what I have done, and I am wondering if I understand correctly that NA include unknown,multiple,and "" in chromosome coloumn
* Good job, on changing the color

### Missing data and amount of heterozygosity and homozygosity
* Looks very nice
* The group on x-axis is a little hard to see
* Brillant ideas on observation of all the groups in Group column not only just maize and teosinte!

### Own visualization
* Nice figure overall! I like the pink color

!!**Overall very nice summarization**!!

*End of the review session*



# Valerie's R Assignment part and result 
Setting up the R Markdown and loading in R library.

```
rm(list = ls())
library(tidyverse)
library(ggplot2)
library(tidyr)
library(tibble)
library(dplyr)
```


## Part I
### _Data Inspection_
`fang_et_al_genotypes.txt`

* There are 2782 rows and 986 columns.
* The first 3 columns are `Sample_ID`, `JG_OTU`, and `Group`. Then the rest are SNP data (983).

`snp_position.txt`

* There are 983 rows and 15 columns.
* The columns are named `SNP_ID`, `Chromosome`, `Position`, etc.

```{r}
# Loading in data from "data" folder and looking at column names. 
genotypes <- read_tsv("data/fang_et_al_genotypes.txt")
snp_position <- read_tsv("data/snp_position.txt")


# Inspecting the data
dim(genotypes) # 2782 rows  986 columns
colnames(genotypes)[1:15] # variable names 
# head(genotypes) # 1st Sample_ID, 3rd Group
table(genotypes$Group) 
  # maize count: ZMMIL 290, ZMMLR 1256, and ZMMMR 27 = 1573
  # teosinte count: ZMPBA 900, ZMPIL 41, and ZMPJA 34 = 975


dim(snp_position) # 983 rows  15 columns
colnames(snp_position)[1:15] # variable names
# head(snp_position) # 1st SNP_ID, 3rd Chromosome, 4th Position, etc.
table(snp_position$Chromosome) # 6 multiple and 27 unknown
```

### _Data Processing_
`fang_et_al_genotypes.txt` 
First, I am cutting out the groups I need for maize and teosinte.
```{r}
# Filtering `fang_et_al_genotypes.txt` 
maize_genotypes <- genotypes %>% filter(Group == "ZMMIL" |
                                        Group == "ZMMLR" |
                                        Group == "ZMMMR" )
teosinte_genotypes <- genotypes %>% filter(Group == "ZMPBA" |
                                           Group == "ZMPIL" |
                                           Group == "ZMPJA" )
# Checking: both result is 3, which is accurate
n_distinct(maize_genotypes$Group)
n_distinct(teosinte_genotypes$Group) 

```

Then I will transpose the genotype files so that columns become rows by using t() before joining.
```{r}
# MAIZE
t_maize_genotypes <- as.data.frame(t(maize_genotypes))
t_maize_genotypes <- t_maize_genotypes[-c(1,2,3),]
dim(t_maize_genotypes) # 983 rows just like snp_position
# Add Sample_ID back
colnames(t_maize_genotypes) <- maize_genotypes$Sample_ID

# TEOSINTE
t_teosinte_genotypes <- as.data.frame(t(teosinte_genotypes))
t_teosinte_genotypes <- t_teosinte_genotypes[-c(1,2,3),]
dim(t_teosinte_genotypes) # 983 rows just like snp_position
# Add Sample_ID back
colnames(t_teosinte_genotypes) <- teosinte_genotypes$Sample_ID


```

Now I am cleaning snp_position and fixing the join key for genotype files.
Assignment says files formated with 1st column is "SNP_ID", the 2nd column is "Chromosome", the 3rd column is "Position", and subsequent columns are genotype data from either maize or teosinte individuals."
_https://stackoverflow.com/questions/29511215/how-can-i-convert-row-names-into-the-first-column_ 
```{r}
# Clean snp_position. We only want column 1, 3, 4
cleaned_snp_position <- select(snp_position, 
                               "SNP_ID", "Chromosome", "Position")
# I don't have SNP_ID column name from genotype data so here I add it.
t_maize_genotypes <- tibble::rownames_to_column(t_maize_genotypes, "SNP_ID")
t_teosinte_genotypes <- tibble::rownames_to_column(t_teosinte_genotypes, "SNP_ID")

```

Joining snp_position to both of these genotype files.
_https://r4ds.hadley.nz/joins.html_
```{r}
# MAIZE
joined_maize <- left_join(
  x = t_maize_genotypes, 
  y = cleaned_snp_position, 
  by = "SNP_ID")
tail(names(joined_maize), 5) # it stuck it onto the end, lets move it to the front
joined_maize <- joined_maize %>% select(SNP_ID, Chromosome, Position, everything())


# TEOSINTE
joined_teosinte <- left_join(
  x = t_teosinte_genotypes, 
  y = cleaned_snp_position, 
  by = "SNP_ID")
tail(names(joined_teosinte), 5) # it stuck it onto the end, lets move it to the front
joined_teosinte <- joined_teosinte %>% select(SNP_ID, Chromosome, Position, everything())

```

Now let's check the new datasets : `joined_maize` and `joined_teosinte`
```{r}
unique(joined_maize$Chromosome) # we need to remove multiple and unknown
joined_maize2 <- filter(joined_maize, 
                Chromosome != "multiple" & Chromosome != "unknown")
unique(joined_maize2$Chromosome) # gone!

unique(joined_teosinte$Chromosome) # we need to remove multiple and unknown
joined_teosinte2 <- filter(joined_teosinte, 
                Chromosome != "multiple" & Chromosome != "unknown")
unique(joined_teosinte2$Chromosome)  # gone!
```

Fixing position and removing positions that are not numeric, and hence can not be sorted by increasing position values.
```{r}
class(joined_maize2$Position) # character, should be numeric so I don't sort wrong
cleaned_maize <- joined_maize2 %>%
  mutate(Position = as.numeric(Position)) %>%
  filter(!is.na(Position))

cleaned_teosinte <- joined_teosinte2 %>%
  mutate(Position = as.numeric(Position)) %>%
  filter(!is.na(Position))

```
This is how I would generate the final files for 1 chromosome.
```{r}

dim(cleaned_maize) # 1576 columns
test_chr_1 <- cleaned_maize %>%
  filter(Chromosome == "1")
test_chr_1_increase <- test_chr_1 %>%
  arrange(Position) %>%
  mutate(across(4:1576, function(x){
    ifelse(x == "?/?", "?", x)
  })
  )

test_chr_1_decrease <- test_chr_1 %>%
  arrange(desc(Position))%>%
  mutate(across(4:1576, function(x){
    ifelse(x == "?/?", "-", x)
  })
  )
``` 

Creating a function to do all 10!
```{r}

final_files <- function (df, chr, name){
  
  chr_df <- df %>% filter(Chromosome == chr)
  chr_increase <- chr_df %>% 
    arrange(Position) %>%
    mutate(across(-(1:3), function(x){
    ifelse(x == "?/?", "?", x)
  })
  )
  
  chr_decrease <- chr_df %>%
  arrange(desc(Position))%>%
  mutate(across(-(1:3), function(x){
    ifelse(x == "?/?", "-", x)
  })
  )
# write_csv(dataframe, output path should be final_results, delimiters) 
# I need no spaces with it in the final_maize or final_teosinte folder which is inside the final_results folder. 
# set up should be : final_results/final_maize/maize_chr1_increasing.csv
  # which means that the actual file names requires 
    # name + _chr + "chr" (the actual value) + _increasg.csv
   write_csv(chr_increase, paste0("final_results/final_", name, "/", 
                                  name, "_chr", chr, 
                                  "_increasing.csv") )
   write_csv(chr_decrease, paste0("final_results/final_", name, "/", 
                                  name, "_chr", chr, 
                                  "_decreasing.csv") )

}

```

Using `for loop` since it's fairly short?
```{r}
class(cleaned_maize$Chromosome) # characters

for(chr in as.character(1:10)) {
  final_files(cleaned_maize, chr, "maize") 
}

for(chr in as.character(1:10)) {
  final_files(cleaned_teosinte, chr, "teosinte") 
}
```

## Part II Visualization

### _SNPs per chromosome_
__What is the distribution of SNPs on and across chromosomes?__

The graph below answers this question.
From most to least number of SNPS are:
Chromosome 1, 2, 5, 3, 7, 4, 6, 8, 9, 10.

```{r}
# Unsure if we want Chromosome with multiple or unknown since they aren't numerically sortable.
# This gives us number of SNP sites on each chromosome 1 through 10.

dist_SNPs <- snp_position %>% 
  filter(Chromosome != "multiple" & Chromosome != "unknown") %>%
  mutate(Chromosome = as.numeric(Chromosome)) %>%
  mutate(Chromosome = factor(Chromosome, levels = 1:10)) %>%
  count(Chromosome)
 

  
ggplot(dist_SNPs, aes(x=Chromosome, y=n)) +
  geom_bar(stat = "identity", fill = "lightblue2") + 
  labs (
    title = "Distribution of SNPs on and across chromosomes",
    x = "Chromosome",
    y = "Number of SNPs"
  ) 
```

__Are there more SNP positions in maize or teosinte individuals?__ 

There are more SNP positions in __teosinte__ individuals (there are more polymorphic SNP sites).

I believe I am looking for polymporphic SNPS. My thought process is to use `joined_maize` before I removed multiple and unknown chromosomes and non-numeric positions since we want to look at maize and teosinte as a whole.
```{r}
poly_maize <- apply(joined_maize[, -(1:3)],1, 
# we don't want the first 3 columns since that's not genotype data, and so given the data we want, for every row x we do...
              function (x) {
                x <- x[x != "?/?"]
# I don't want missing data because it gets in the way of sorting by unique
                length(unique(x)) > 1 })
# Checking if variation, if all the same the monomorphic
# view(poly_maize) # True for polymorphic, False for monomorphic!

joined_maize$polymorphic <- poly_maize
# Tacking on a column to end of joined_maize2

poly_maize <- joined_maize  %>% 
  filter(polymorphic)  %>% 
  count(Chromosome)  %>% 
  mutate(Dataset = "maize")

# exact same thing for teosinte

poly_teosinte <- apply(joined_teosinte[, -(1:3)],1, 
              function (x) {
                x <- x[x != "?/?"]
                length(unique(x)) > 1 })
joined_teosinte$polymorphic <- poly_teosinte
# Tacking on a column to end of joined_maize

poly_teosinte <- joined_teosinte  %>% 
  filter(polymorphic)  %>% 
  count(Chromosome)  %>% 
  mutate(Dataset = "teosinte")

```

Plot `teosinte` and `maize` now!
```{r}
poly_per_chr <- bind_rows(poly_maize, poly_teosinte)
#view(poly_per_chr)

poly_per_chr <- mutate(poly_per_chr, Chromosome = factor(Chromosome, levels = as.character(1:10)))

ggplot(poly_per_chr, aes(x = Chromosome, y = n, fill = Dataset)) + geom_bar(stat = "identity", position = "dodge") + labs (
      title = "Polymorphic SNP positions: chromosomes 1 - 10",
      x = "chromosome",
      y = "# of polymorphic SNPs"
)  + scale_fill_manual(values = c("maize" = "goldenrod1", "teosinte" = "forestgreen"))

```

### _Missing data and amount of heterozygosity_
__What is the proportion of homozygous and heterozygous sites as well as missing data in each sample and each group?__

```{r}
# here I am sorting for missing, homozygous, and heterzygous by using substr() to see if the genotypes match and sum counts how many sites
homhetmis <- genotypes %>%
  mutate(
    missing = apply(select(., -(1:3)), 1, 
                    function(x) sum(x == "?/?")),
    homozygous = apply(select(., -(1:3)), 1, 
                    function(x) sum(x != "?/?" & substr(x,1,1) == substr(x,3,3))),
    heterozygous = apply(select(., -(1:3)), 1, 
                    function(x) sum(x != "?/?" & substr(x,1,1) != substr(x,3,3)))
    )

# now I'm totaling the sites
homhetmis <- homhetmis %>%
  mutate(total = missing + homozygous + heterozygous,
         proportion_miss = missing / total,
         proportion_homo = homozygous / total,
         proportion_hete = heterozygous / total)

# now I need to check my dataframe and take what I need
select(homhetmis, Sample_ID, Group, proportion_miss, proportion_hete, proportion_homo)

# now putting Samples into their Group
group_homhetmis <- homhetmis %>%
  group_by(Group) %>%
  summarise(
    mean_miss = mean(proportion_miss),
    mean_hete = mean(proportion_hete),
    mean_homo = mean(proportion_homo)
  )
# creating that new column to indicate whether a particular site is homozygous
plot_homhetmis <- bind_rows(
  group_homhetmis %>%
    select(Group, prop = mean_miss) %>%
  mutate(site = "missing"),

  group_homhetmis %>%
   select(Group, prop = mean_homo) %>%
  mutate(site = "homozygous"),   
 
   group_homhetmis %>%
     select(Group, prop = mean_hete) %>%
  mutate(site = "heterozygous"))

ggplot(plot_homhetmis, aes(x = Group, y = prop, fill = site)) +
  geom_col(position = "fill") + labs(
    title = "Proportion of missing, homozygous, and heterozygous sites by group",
    x = "group",
    y = "propotion"
  ) +
#?scale_fill_manual
  scale_fill_manual(values = c("missing" = "mediumpurple2", "homozygous" = "lightblue2", "heterozygous" = "coral2"))


```

### _Your own visualization_
__Visualize one other feature of the dataset. The choice is up to you!__
```{r}
# I am plotting the genomoic position of polymorphic sites.
# Since I used joined_maize for the polymorphic work above, I'll use it again. 

plot_mazie_poly <- joined_maize  %>% 
  filter(polymorphic)  %>% 
  
# I have to remove unknown and multiples again, they created an N/A column  
  mutate(Dataset = "maize", Position = as.numeric(Position) ) %>%
  filter(!is.na(Position)) %>%

# Sorts alphanumerically again so....
  mutate(Chromosome = factor(Chromosome, levels = 1:10))
  

plot_teosinte_poly <- joined_teosinte  %>% 
  filter(polymorphic)  %>% 
  mutate(Dataset = "teosinte", Position = as.numeric(Position) ) %>%
  filter(!is.na(Position)) %>%
  mutate(Chromosome = factor(Chromosome, levels = 1:10))

# combine them
plot_poly_position <- bind_rows(plot_mazie_poly, plot_teosinte_poly )

# increased binwidth, small numbers spit error
ggplot(plot_poly_position, aes(x=Position)) +
  geom_histogram(binwidth=1000000, fill = "black", color = "hotpink2") + 
  facet_wrap( ~ Chromosome) + 
  
  labs (
    title = "Distribution of SNP position across chromosomes 1-10",
    x = "position",
    y = "# of SNPs"
  ) 
```

